Infinite products and zero--one laws in categorical probability

### 2020-05-10 response_to_referees.tex

\renewcommand{\sf}{\mathsf}
\DeclareMathOperator{\del}{del}
\begin{document}
\maketitle

\begin{abstract}
	We would like to thank both referees for their useful suggestions. There are some items that we would like to respond to, in particular from the first referee's report.
\end{abstract}

\section*{Referee 1}

We have been quite surprised by the length and the amount of detail of this report, and it is exciting that the referee has invested so much work into it. The referee's suggestion of putting more emphasis on our infinite tensor products and Kolmogorov products has been helpful for us, and we have shifted the emphasis a bit.

At the same time, we strongly disagree with the referee on essentially all other counts. In particular, the details below will explain that all ``errors'' pointed out in the report are the referee's errors, not ours. We start by commenting on the summary report and then move on to the material in the attached PDF.

@@

		In fact, the referee already seems to agree with this: the relevant diagram is exactly the referee's one on the second half of p.6 (in the report). The ``new terminology'' that the referee proposes surely does not have any bearing on whether the universal property described by the diagram holds in $\BorelStoch$, right?
\end{enumerate}

On the comments in the PDF:

\begin{enumerate}[resume,label=(\alph*).]
	\item The referee's proposed new formulation of Definition 3.1 is along the lines of our Definition 4.1.\footnote{With the only difference between the preservation by the functors $-\otimes Y$, which we get to in the next item.} In both definitions, the Kolmogorov product is the limit of the finite products $\bigotimes_{i \in F} X_i$ in both $\cC$ and in $\cC_{\det}$. So it is not clear to us what the referee's point is.

		Perhaps the referee's suggestion is to replace Definition 3.1 by Definition 4.1. We do not want to do this, since the latter only applies in Markov categories, while the former more generally makes sense in semicartesian monoidal categories.
		\begin{displayquote}
			and hence every ``infinite tensor product'' is Kolmogorov.
		\end{displayquote}
		This is true in categories like $\Stoch$ and $\BorelStoch$, but not in general. We had already explained this point after Definition 4.1. In conclusion, the distinction between infinite tensor products (in semicartesian monoidal categories, Section 3) and Kolmogorov products (in Markov categories, Section 4) is important and should not be swept under the rug.
	\item Quoting from p.7,
		\begin{displayquote}
			I did not incorporate the property of preservation of lim $\mathcal{D}$ by the endofunctor
			$-\otimes Y : \cC_{\det} \to \cC_{\det}$ because that condition is a global property in the sense
		that $Y$ is an arbitrary element in your space which may have no relationship
		to your fixed family $(X_j)_{j\in J}$ for which you are trying to find an extension.
		\end{displayquote}
		Assuming that ``space'' means category, we agree that there is no relation between $Y$ and the family $(X_i)$, but we do not see why and how this would be an argument against our preservation condition. As the referee will hopefully have noticed, this preservation condition is an important property which we use throughout the paper.

		Perhaps the referee's point is that this preservation condition never appears in the context of Kolmogorov's extension theorem. This is true, and the reason is simple: it \emph{automatically holds} in $\BorelStoch$, as we have explained in Example 3.6. Hence there is no reason to state it explicitly in the standard probability context.
	\item On p.8, the referee again seems to conclude the incorrectness of a statement from their lack of understanding of our arguments:
		\begin{displayquote}
		(2) Page 17, the last sentence. The statement
			\begin{displayquote}
			Moreover, $\hat{\sigma}$ is deterministic by Proposition 4.3.
			\end{displayquote}
		is not correct.
		\end{displayquote}
		The comment that follows:
			\begin{displayquote}
				then please cite a reference	
			\end{displayquote}
		indeed makes it clear that the referee does not claim to have found a mistake. In fact, the reason that $\hat{\sigma}$ is deterministic is by a trivial application of the universal property of the Kolmogorov product as a limit in $\cC_{\det}$, based precisely on the facts that the referee mentions: both the composite and the $\pi_F$'s are deterministic, and hence the factorization exists uniquely.
	\item Again p.8:
		\begin{displayquote}
			(3) Corollary 6.1. (Kolmogorov zero-one law) Your assumption, which says that
			$T \subseteq \Omega$, is incorrect, because if $T$ is just a subset of $\Omega$, then I can just choose
			it to be any nonmeasurable set of $\Omega$.
		\end{displayquote}
		It is a mystery to us how it is possible to read over half of a sentence: the relevant measurability assumption was stated in clear and explicit terms directly after $T \subseteq \Omega$.
		\begin{displayquote}
			If you
			meant to say that $T$ is a measurable set in $\Omega$, the whole problem statement is an
			elementary result in measure theory, and certainly not the Kolmogorov zero-one
			law.
		\end{displayquote}
		If $T$ was an \emph{arbitrary} measurable set in $\Omega$, then the statement would quite obviously be false. In any case, our statement of the Kolmogorov zero--one law has been and is exactly the classical one, and it is so classical that it should not require a reference.
	\item p.9, item (2): Even if most categorical foundations in algebra and geometry required monoidal closure (which is obviously not the case, take e.g.~abelian categories), we do not see how this would be an argument for monoidal closure in the context of probability theory.
	\item p.9, item (4): Standard usage of the term ``diagram'' in category theory does not have any association with graph homomorphisms. In standard parlance as we use it, ``diagram'' is just any functor from a small category into the category under consideration.\footnote{See e.g.~\href{https://ncatlab.org/nlab/show/diagram\#CategoryShapedDiagramFunctorially}{https://ncatlab.org/nlab/show/diagram\#CategoryShapedDiagramFunctorially}.}
	\item p.11:
		\begin{displayquote}
			\textbf{Theorem 4.1.} The weak limit of the functor $\mathcal{D}' : \mathscr{J} \to \Stoch$ exists. Moreover, the weak limit of the functor $\mathcal{D}'$ is never a limit in $\Stoch$ except for trivial cases.
		\end{displayquote}
		It is not clear to us what ``the weak limit'' means, since weak limits are rarely unique (up to unique iso), and in particular uniqueness does not hold in the case under consideration. That being said, we agree that the referee's construction indeed produces a weak limit that is not a limit. But the fact that that particular weak limit is not a limit says nothing about whether the functor has a limit or not.
\end{enumerate}

@@

\section*{Referee 2}

We thank the referee for a number of suggestions which have led to an improvement of the presentation. It is good to know for us that the paper is not so easy to follow at times, and we have implemented . While an exposition of string diagrams and Markov categories would perhaps be a bit too much for this paper, we hope that these shortcomings can be addressed in a possible future textbook treatment of the subject.

### 2020-05-10 categorical-zero-one.tex
some revision

\author{Tobias Fritz and Eigil Fjeldgren Rischel}
\title{The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability}
\date{\today}

@@

\begin{abstract}
	We state and prove the zero--one laws of Kolmogorov and Hewitt--Savage within the setting of \emph{Markov categories}, a category-theoretic approach to the foundations of probability and statistics. This gives general versions of these results which can be instantiated not only in measure-theoretic probability, where they specialize to the standard ones in the setting of standard Borel spaces, but also in other contexts. For example, applying the Kolmogorov law to the Kleisli category of the hyperspace monad on topological spaces gives criteria for when maps out of an infinite product of topological spaces into a Hausdorff space are constant.
\end{abstract}

@@

\emph{Markov categories} are an approach to the foundations of probability and statistics based on category theory, proposed first by Golubtsov~\cite{golubtsov}, rediscovered recently independently by Cho and Jacobs~\cite{cho_jacobs}, and developed extensively by the first author in \cite{markov_cats}. The basic observation is that Markov kernels can be composed sequentially and in parallel, making them into a symmetric monoidal category $\Stoch$. One can then find axioms which make a symmetric monoidal category behave similarly to the actual category of Markov kernels, and then state and prove results from probability and statistics in this general abstract context.

This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}). Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry: here, points and lines are described only in terms of their relationships with one another; while in the analytic approach often attributed to Descartes, geometric objects are defined concretely in terms of more primitive notions like sets and real numbers.

In this paper, we give a treatment of the zero--one laws of Kolmogorov and of Hewitt--Savage in the setting of Markov categories. This adds further support to the hypothesis that Markov categories are an adequate setting for the foundations of probability and statistics, adding to the existing treatments of Bayesian updating, almost sure equality, conditional independence, and theorems on sufficient statistics~\cite{cho_jacobs,markov_cats}.

Very roughly, the Kolmogorov zero--one law states that, given an infinite collection of independent random variables and an event which is in the $\sigma$-algebra generated by the variables but independent of any finite subset of these variables, the event has probability zero or one.
This theorem goes back to Kolmogorov's foundational monograph \cite{kolmogorovgerm}, although the statement appearing there is significantly different from the modern form.
For a more recent treatment of this theorem, see e.g. \cite[Theorem~2.37]{klenke}

The Hewitt--Savage zero--one law, due to Hewitt and Savage in \cite[Theorem~11.3]{hewitt_savage}, is similar. Here, the variables are assumed to be identically distributed in addition to being independent, and the event is assumed to be invariant under finite permutations of the variables. Again, under these assumptions, the event has probability zero or one.

To formulate these statements in the general context of Markov categories, we need an abstract generalization of the concept of joint distribution of infinitely many random variables.
To this end, we introduce a notion of \emph{infinite tensor product}, first for semicartesian monoidal categories (\Cref{infprod_semicartesian}) and then for Markov categories (\Cref{infprod_markov}). A morphism to such an infinite tensor product is an abstract version of a probability measure on an infinite product of measurable spaces. Based on this definition, we then prove abstract versions of the zero--one laws, \Cref{thm:kolmog} and \Cref{thm:hewsav}.

Finally in \Cref{examples}, we apply our results to the category of measurable spaces and Markov kernels $\Stoch$. In order to guarantee that countably infinite tensor products exist, we restrict to the full subcategory of standard Borel spaces. We thereby rederive the standard zero--one laws for random variables in standard Borel spaces from our abstract theorems. We also instantiate one of our theorems in the Kleisli category of the lower Vietoris monad, which gives criteria for when a continuous map out of infinite product of topological spaces into a Hausdorff space is constant (\Cref{vietoris_kolmogorov}).

@@

Throughout, $\cC$ is a symmetric monoidal category. To simplify notation, we generally omit mention of the structure isomorphisms by assuming that $\cC$ is strict without loss of generality. We make frequent use of string diagram notation, and in doing so we omit object labels whenever these are obvious from the context.

@@

A symmetric monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently~\cite[Theorem~3.5]{GLS}, it is a symmetric monoidal category equipped with morphisms
\begin{equation}
	\label{projections}
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\end{equation}
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. Here are the main examples that we will be considering.

\begin{example}
	\label{finstoch}
	$\FinStoch$ is the category of finite sets with \emph{stochastic matrices}, which means that a morphism $f : X \to Y$ is a matrix $(f_{xy})_{x \in X,y \in Y}$ of nonnegative real numbers such that $\sum_y f_{xy} = 1$ for every $x$. Stochastic matrices compose by matrix multiplication, which in this context is also known as the \emph{Chapman--Kolmogorov equation}. We consider $\FinStoch$ as a symmetric monoidal category with respect to the cartesian product of sets on objects and the Kronecker product of stochastic matrices on morphisms; the monoidal structure isomorphisms are the obvious ones coming from the full embedding $\FinSet \subseteq \FinStoch$.
\end{example}

@@

\begin{example}
	\label{stoch}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. A \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a map
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, one can think of $f$ as assigning to every $x \in X$ a random element of $Y$. Composition is again defined by a variant of the Chapman--Kolmogorov equation, and the symmetric monoidal structure is defined in terms of the usual product of measurable spaces. The monoidal unit $I$ is given by any one-element set with its unique $\sigma$-algebra. As a special case, morphisms $I \to (X,\Sigma_X)$ can be identified with probability measures $\Sigma_X \to [0,1]$.

	There is a well-known subclass of particularly well-behaved measurable spaces, the \emph{standard Borel spaces}, or equivalently Polish spaces equipped with their Borel $\sigma$-algebras. We write $\BorelStoch \subseteq \Stoch$ for the full subcategory of standard Borel spaces with Markov kernels. Since finite products of standard Borel spaces are again standard Borel, $\BorelStoch$ is again semicartesian symmetric monoidal.
\end{example}

@@

Our goal is to use semicartesian monoidal categories in order to develop aspects of probability theory in categorical terms, in such a way that instantiating this theory in $\Stoch$ or $\BorelStoch$ recovers the standard theory. As it turns out, doing so requires a bit more structure, in a form which has been axiomatized first by Cho and Jacobs~\cite{cho_jacobs} as \emph{affine CD-categories}, although very similar definitions occur in earlier work of Golubtsov~\cite{golubtsov}. We here follow the terminology of our own~\cite[Definition~2.1]{markov_cats}.

@@

There is a strictification result~\cite[Theorem~10.17]{markov_cats} which guarantees that $\cC$ can be assumed to be strict monoidal, as we do throughout.

@@

\Cref{finstoch,cring,stoch} are all Markov categories in a canonical way. In $\FinStoch$ and $\Stoch$, the comultiplication morphisms are given indeed by copying, i.e.~by those stochastic matrices or Markov kernels which map an $x \in X$ to the Dirac delta measure at $(x,x) \in X \times X$. We refer to~\cite[Example~2.5 and Section~4]{markov_cats} for the technical details. In $\CRing_+\op$, we define the comultiplication of an object $R \in \CRing_+\op$ to be represented by the multiplication map $R \otimes R \to R$. The multiplicativity condition~\eqref{multiplicativity} then amounts to the fact that the multiplication on a tensor product of commutative rings $R \otimes S$ is given by the defining equation
\[
	(r_1 \otimes s_1) (r_2 \otimes s_2) \, = \, r_1 r_2 \otimes s_1 s_2.
\]

@@

In $\FinStoch$, the deterministic morphisms $X \to Y$ are exactly the $\{0,1\}$-valued stochastic matrices. Since these can be identified with honest functions $X \to Y$, we conclude that $\FinStoch_\detc$ is equivalent to $\FinSet$. In $\Stoch$, the deterministic morphisms $f : X \to Y$ are exactly those Markov kernels for which $f(S|x) \in \{0,1\}$ for every $S \in \Sigma_Y$ and $x \in X$. In other words, as the term suggests, the deterministic morphisms are those which do not involve any randomness; in $\BorelStoch$, the deterministic morphisms can be identified with the measurable functions. In $\CRing_+\op$, the deterministic morphisms $R \to S$ are precisely those morphisms which are represented by additive unital maps $S \to R$ which are in addition multiplicative, i.e.~the ring homomorphisms. Thus the subcategory of deterministic morphisms is exactly the opposite of the usual category of commutative rings.

We state some more definitions that we will be using later on.

@@

It is known that $\Stoch$ is causal~\cite[Example~11.35]{markov_cats}, and
therefore so are the subcategories $\BorelStoch$ and $\FinStoch$.

\begin{example}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[t]$, where the following counterexample similar to~\cite[Example~11.33]{markov_cats} can be formulated.
Consider the additive unital maps given as the additive extensions of
\begin{align*}
	f(t^{n}) = {}& \begin{cases}t^{n-1} & n \geq 1 \\ 1 & n = 0 \end{cases}\\[3pt]
	g(t^{n}) = {}& \begin{cases}t & n \geq 1 \\ 1 & n= 0\end{cases}\\[3pt]
	h_{1}(t^{n}) = {}& t^{n}\\[3pt]
	h_{2}(t^{n}) = {}& 1
\end{align*}
On the basis monomials $t^n$, we have $(fg)(t^n) = 1$, and therefore
\[
	(fg)(h_1(t^n)t^m) = 1 = (fg)(h_2(t^n)t^m)
\]
for all $n,m \in \mathbb{N}$, so that the hypothesis holds.
If $\CRing_{+}^{\op}$ were causal, then we would also have
\[
	f(g(h_{1}(t^n) t^m) t^\ell) = f(g(h_{2}(t^n) t^m) t^\ell)
\]
for all $n,m,\ell \in \mathbb{N}$. But this clearly fails for $n = \ell = 1$ and $m = 0$.
\end{example}

@@

  This term comes from measure theory, and should be interpreted as meaning
  that for all those values of $X$ which can occur as outputs of $p$, feeding these values to $f$ and $g$ as inputs is guaranteed to result in the same distribution; at the same time, the behavior of $f$ and $g$ may be different on those input values which do not occur as outputs of $p$.
  \todo[inline]{I've rephrased this in order to make sure that we don't create the impression of making a technically precise statement}

We finally introduce a straightforward generalization of~\cite[Defnition~12.12]{markov_cats} from binary conditional independence to $n$-ary conditional independence.

@@

Zero--one laws involve infinitely many random variables. In the setting of Markov categories, this means that we need to consider morphisms whose codomain is an ``infinite tensor product'' in a suitable sense. In this section, we start by considering infinite tensor products in semicartesian symmetric monoidal categories.

Suppose that $(X_i)_{i \in J}$ is any family of objects, where the indexing set $J$ is typically infinite. For any finite subset $F \subseteq J$, we also write $X_F := \bigotimes_{i \in F} X_i$ for simplicity of notation. In the spirit of Kolmogorov's extension theorem, we believe that a good definition of joint distribution of infinitely many random variables will be such that such a joint distribution is uniquely determined by the family of its finite marginals. This motivates our general definition of the infinite tensor product, which we now introduce. If $F \subseteq F' \subseteq J$ are two finite subsets, then the fact that $\cC$ is semicartesian monoidal gives us \emph{marginalization morphisms}
\[
	\pi_{F',F} \: : \: X_{F'} \longrightarrow X_F.
\]
Via these maps, the finite tensor products $(X_F)_{F \subseteq J}$ make up a cofiltered diagram in the form of a functor from the poset of finite subsets of $J$, ordered by reverse inclusion, to $\cC$.

>>>>

\author{Tobias Fritz and Eigil Fjeldgren Rischel}
\title{Infinite products and zero--one laws in categorical probability}
\date{\today}

@@

\begin{abstract}
	\emph{Markov categories} are a recent category-theoretic approach to the foundations of probability and statistics. Here we develop this approach further by treating infinite products and the Kolmogorov extension theorem. This is relevant for all aspects of probability theory in which infinitely many random variables appear at a time. These infinite tensor products $\bigotimes_{i \in J} X_i$ come in two versions: a weaker but more general one for families of objects $(X_i)_{i \in J}$ in semicartesian symmetric monoidal categories, and a stronger but more specific one for families of objects in Markov categories.

	As a first application, we state and prove versions of the zero--laws of Kolmogorov and Hewitt--Savage laws for Markov categories. This gives general versions of these results which can be instantiated not only in measure-theoretic probability, where they specialize to the standard ones in the setting of standard Borel spaces, but also in other contexts. % For example, applying the Kolmogorov zero--one law to the Kleisli category of the hyperspace monad on topological spaces gives a criterion for when maps out of an infinite product of topological spaces into a Hausdorff space are constant.

\end{abstract}

@@

\emph{Markov categories} are an approach to the foundations of probability and statistics based on category theory, proposed first by Golubtsov~\cite{golubtsov}, rediscovered recently independently by Cho and Jacobs~\cite{cho_jacobs}, and developed extensively by the first-named author in \cite{markov_cats}. The basic observation is that Markov kernels can be composed sequentially and in parallel, making them into a symmetric monoidal category $\Stoch$. One can then find axioms which make a symmetric monoidal category behave similarly to the actual category of Markov kernels, and then state and prove results from probability and statistics in this general abstract context. Thus Markov categories are abstract versions of the category of Markov kernels, and hence their name.

This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}). Instead of defining \emph{what} the words probability, distribution, random variable, etc, mean semantically, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy with respect to each other.
We can draw an analogy to the Euclidean approach to geometry: in this synthetic approach, points and lines are described only in terms of their relationships with one another. This is to be contrasted with the analytic approach often attributed to Descartes, in which geometric objects are defined concretely in terms of more primitive notions like sets and real numbers.

In this paper, we give a treatment of further fundamental and classical results of probability theory within the setting of Markov categories. This adds further support to the hypothesis that Markov categories are an adequate setting for the foundations of probability and statistics, adding to the existing treatments of Bayesian updating, almost sure equality, conditional independence, and theorems on sufficient statistics~\cite{cho_jacobs,markov_cats}. Concretely, we develop a notion of infinite tensor products in Markov categories and then apply this notion to state and prove two of the standard zero--one laws of probability theory.

\paragraph*{Summary.}

We now present a more detailed overview.

\Cref{background} presents some background by recalling the definition of Markov category, presenting some pertinent examples, and discussing those aspects of the existing theory of Markov categories that will be of relevance for the rest of the paper. With the exception of \Cref{cring_not_causal}, there is nothing new in this section.

\Cref{infprod_semicartesian} then proceeds by introducing infinite tensor products in semicartesian symmetric monoidal categories, which are exactly those symmetric monoidal categories in which the monoidal unit object is terminal. We consider a number of examples of such categories and discuss whether infinite tensor products exist. This already includes examples relevant for probability theory, and in particular $\BorelStoch$, the category of standard Borel spaces and measurable maps\footnote{Or equivalently Polish spaces and measurable maps.}, for which we explain how countably infinite tensor products implement the Kolmogorov extension theorem (for countably many variables); the extension theorem states that the joint distributions of infinitely many random variables are in bijection with compatible families of joint distributions for each finite subset of variables. The universal property of our infinite tensor products makes the extension theorem into a definition, generalized from distributions to Markov kernels. An important subtlety here is the additional preservation condition in \Cref{semicartesian_infproduct}, where infinite tensor products are introduced: this additional condition guarantees that infinite tensor products have nice compositionality properties, so that e.g.~a tensor product of two infinite tensor products is itself an infinite tensor product (\Cref{two_infproducts}). This condition does not come up traditionally in the context of the Kolmogorov extension theorem, the reason being that it automatically holds in the case of $\BorelStoch$ (\Cref{infprods_stoch}).

In the short \Cref{infprod_markov}, we then turn to infinite tensor products in Markov categories, which we call \emph{Kolmogorov products} due to the connection with the Kolmogorov extension theorem. In addition to being an infinite tensor product, a Kolmogorov product also needs to respect the additional structure which makes a semicartesian symmetric monoidal category into a Markov category, namely the comultiplications $\cop{X} : X \to X \otimes X$. This is implemented by postulating that the product projections should be deterministic morphisms, or equivalently that the infinite product should be an honest categorical product in the cartesian monoidal subcategory of deterministic morphisms (\Cref{kolmprod_catdet}). Not surprisingly, the infinite tensor products in $\BorelStoch$, which implement the Kolmogorov extension theorem, are indeed Kolmogorov products.

In \Cref{zeroones}, we state and prove the zero--one laws of Kolmogorov and Hewitt--Savage in their synthetic versions. Very roughly, the Kolmogorov zero--one law states that, given an infinite collection of independent random variables and an event which is in the $\sigma$-algebra generated by the variables but independent of any finite subset of them, the event has probability zero or one. This theorem goes back to Kolmogorov's foundational monograph \cite{kolmogorovgerm}, although the statement appearing there is significantly different from its modern form. For a more recent textbook treatment of this theorem, see e.g. \cite[Theorem~2.37]{klenke}

The Hewitt--Savage zero--one law, due to Hewitt and Savage in \cite[Theorem~11.3]{hewitt_savage}, is similar in flavour but technically quite different. Here, the infinitely many variables are assumed to be identically distributed in addition to being independent, and the event is assumed to be invariant under finite permutations of the variables. Again, under these assumptions, the event has probability zero or one.

Finally in \Cref{examples}, we apply our results to $\BorelStoch$. We thereby rederive the standard zero--one laws for random variables in standard Borel spaces from our abstract theorems. We also instantiate one of our theorems in the Kleisli category of the lower Vietoris monad, which gives criteria for when a continuous map out of infinite product of topological spaces into a Hausdorff space is constant (\Cref{vietoris_kolmogorov}).

@@

Throughout, $\cC$ is a symmetric monoidal category. To simplify notation, we generally omit mention of the structure isomorphisms by assuming that $\cC$ is strict without loss of generality. We routinely make use of string diagram notation, and in doing so we omit object labels whenever these are obvious from the context. Our string diagrams are to be read from bottom to top.

@@

We assume familiarity with symmetric monoidal categories, up to and including string diagram notation and string diagram calculus for composite morphisms in symmetric monoidal categories.

A symmetric monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently~\cite[Theorem~3.5]{GLS}, $\cC$ is semicartesian if it comes equipped with morphisms
\begin{equation}
	\label{projections}
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\end{equation}
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. In the probability context, these morphisms can be interpreted as marginalizations, or equivalently as discarding the value of the variable $Y$ or $X$.

Here are the main examples of semicartesian symmetric monoidal categories that we will be considering.

\begin{example}
	\label{finstoch}
	$\FinStoch$ is the category of finite sets with \emph{stochastic matrices} as morphisms. This means that for finite sets $X$ and $Y$, a morphism $f : X \to Y$ is a matrix $(f_{xy})_{x \in X,y \in Y}$ of nonnegative real numbers such that $\sum_y f_{xy} = 1$ for every $x$. Stochastic matrices compose by matrix multiplication, which in this context is also known as the \emph{Chapman--Kolmogorov equation}. We consider $\FinStoch$ as a symmetric monoidal category with respect to the cartesian product of sets on objects and the tensor product (Kronecker product) of stochastic matrices on morphisms; the monoidal structure isomorphisms are the obvious ones coming from the embedding $\FinSet \subseteq \FinStoch$.
\end{example}

@@

\begin{example}
	\label{stoch}
	$\Stoch$ is the category of measurable spaces $(X,\Sigma_X)$ as objects and Markov kernels as morphisms. We sketch the definition here and refer to~\cite[Section~4]{markov_cats} and references therein for the details. Given measurable spaces $(X,\Sigma_X)$ and $(Y,\Sigma_Y)$, a \emph{Markov kernel} $(X,\Sigma_X) \to (Y,\Sigma_Y)$ is a map
	\[
		f \: : \: \Sigma_Y \times X \longrightarrow [0,1], \quad (S,x) \longmapsto f(S|x)
	\]
	assigning to every $x \in X$ a probability measure $f(-|x) : \Sigma_Y \to [0,1]$ in such a way that for every $S \in \Sigma_Y$, the function $x \mapsto f(S|x)$ is measurable. Intuitively, one can think of $f$ as assigning to every $x \in X$ a random element of $Y$; or one can think of $f$ as a statistical model, specifying one probability measure $f(-|x)$ for every parameter value $x$.

	Composition of Markov kernels is again defined by a variant of the Chapman--Kolmogorov equation: for $f : (X,\Sigma_X) \to (Y,\Sigma_Y)$ and $g : (Y,\Sigma_Y) \to (Z,\Sigma_Z)$, the composite is given by, for all $x \in X$ and $T \in \Sigma_Z$,
	\[
		(gf)(T|x) := \int_{y \in Y} g(T|y) \, f(dy|x),
	\]
	where the relevant measurability condition in order for this to define a Markov kernel follows by a standard application of the $\pi$-$\lambda$-theorem; associativity of composition uses Fubini's theorem. In this way, we obtain a category $\Stoch$ which is, more or less by definition, isomorphic to the Kleisli category of the Giry monad~\cite{giry}.

	The symmetric monoidal structure on $\Stoch$ is defined in terms of the usual product of measurable spaces, given by the Cartesian product of underlying sets equipped with the product $\sigma$-algebra. The tensor product of morphisms $f : (A,\Sigma_A) \to (X,\Sigma_X)$ and $g : (B,\Sigma_B) \to (Y,\Sigma_Y)$ is given by the unique Markov kernel between the product measurable spaces which satisfies
	\[
		(f \otimes g)(S \times T|a,b) = f(S|a) \, g(T|b) \qquad \forall a \in A,\: b \in B,\: S \in \Sigma_X,\: T \in \Sigma_Y.
	\]
	The monoidal unit $I$ is given by any one-element set with its unique $\sigma$-algebra. As a special case, morphisms $I \to (X,\Sigma_X)$ can be identified with probability measures $\Sigma_X \to [0,1]$.

	There is a well-known subclass of particularly well-behaved measurable spaces, the \emph{standard Borel spaces}, or equivalently Polish spaces equipped with their Borel $\sigma$-algebras. We write $\BorelStoch \subseteq \Stoch$ for the full subcategory of standard Borel spaces with Markov kernels. Since finite products of standard Borel spaces are again standard Borel, $\BorelStoch$ is again semicartesian symmetric monoidal. In summary, $\BorelStoch$ is the category of standard Borel spaces as objects with measurable Markov kernels as morphisms.
\end{example}

@@

Our goal is to use semicartesian monoidal categories in order to develop aspects of probability theory in categorical terms, in such a way that instantiating this theory in $\Stoch$ or $\BorelStoch$ recovers the standard theory. As it turns out, doing so requires a bit more structure, in a form which has been axiomatized first by Cho and Jacobs~\cite{cho_jacobs} as \emph{affine CD-categories}, although very similar definitions occur in earlier work of Golubtsov~\cite{golubtsov}. We here follow the more intuitive terminology of our own~\cite[Definition~2.1]{markov_cats}.

@@

There is a strictification result~\cite[Theorem~10.17]{markov_cats} which guarantees that $\cC$ can be assumed to be strict monoidal without messing up the commutative comonoid structures. We assume from now on that $\cC$ is strict throughout.

@@

\Cref{finstoch,cring,stoch} are all Markov categories in a canonical way. In $\FinStoch$ and $\Stoch$, the comultiplication morphisms are given indeed by copying, i.e.~by those stochastic matrices or Markov kernels which map an element $x \in X$ to the Dirac delta measure at $(x,x) \in X \times X$. We refer to~\cite[Example~2.5 and Section~4]{markov_cats} for the technical details. In $\CRing_+\op$, we define the comultiplication of an object $R \in \CRing_+\op$ to be represented by the multiplication map $R \otimes R \to R$. The multiplicativity condition~\eqref{multiplicativity} then amounts to the fact that the multiplication on a tensor product of commutative rings $R \otimes S$ is given by the defining equation
\[
	(r_1 \otimes s_1) (r_2 \otimes s_2) \, = \, r_1 r_2 \otimes s_1 s_2.
\]

The following definition is among the central notions of the theory of Markov categories.

@@

In $\FinStoch$, the deterministic morphisms $X \to Y$ are exactly the $\{0,1\}$-valued stochastic matrices, which are indeed those which are deterministic in the sense that they do not involve any randomness. Since these stochastic matrices are in obvious bijection with honest functions $X \to Y$, we conclude that $\FinStoch_\detc$ is equivalent to $\FinSet$. In $\Stoch$, the deterministic morphisms $f : X \to Y$ are exactly those Markov kernels for which $f(S|x) \in \{0,1\}$ for every $S \in \Sigma_Y$ and $x \in X$. In other words, as the term suggests, the deterministic morphisms are again those which do not involve any randomness; in $\BorelStoch$, the deterministic morphisms can be identified with the measurable functions. In $\CRing_+\op$, the deterministic morphisms $R \to S$ are precisely those morphisms which are represented by additive unital maps $S \to R$ which are in addition multiplicative, i.e.~the ring homomorphisms. Thus the subcategory of deterministic morphisms is exactly the opposite of the usual category of commutative rings.

We recall some more definitions that we will be using later on.

@@

Intuitively, this condition states that if the choice between $h_1$ and $h_2$ is irrelevant for what happens ``in the future of $g$'', then that choice is likewise irrelevant for what happened ``in the past of $g$''.
It is known that $\Stoch$ is causal~\cite[Example~11.35]{markov_cats}, and
therefore so are the subcategories $\BorelStoch$ and $\FinStoch$.

\begin{example}
\label{cring_not_causal}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[t]$, where the following counterexample similar to~\cite[Example~11.33]{markov_cats} can be formulated.
Consider the additive unital maps given as the additive extensions of
\begin{align*}
	f(t^{n}) = {}& \begin{cases}t^{n-1} & n \geq 1 \\ 1 & n = 0 \end{cases}\\[3pt]
	g(t^{n}) = {}& \begin{cases}t & n \geq 1 \\ 1 & n= 0\end{cases}\\[3pt]
	h_{1}(t^{n}) = {}& t^{n}\\[3pt]
	h_{2}(t^{n}) = {}& 1
\end{align*}
On the basis monomials $t^n$, we have $(fg)(t^n) = 1$, and therefore
\[
	(fg)(h_1(t^n)t^m) = 1 = (fg)(h_2(t^n)t^m)
\]
for all $n,m \in \mathbb{N}$, so that the hypothesis holds.
If $\CRing_{+}^{\op}$ were causal, then we would also have
\[
	f(g(h_{1}(t^n) t^m) t^\ell) = f(g(h_{2}(t^n) t^m) t^\ell)
\]
for all $n,m,\ell \in \mathbb{N}$. But this clearly fails for $n = \ell = 1$ and $m = 0$.
\end{example}

@@

  This term comes from measure theory, and should be interpreted as meaning
  that for all those values of $X$ which can occur as outputs of $p$, feeding these values to $f$ and $g$ as inputs is guaranteed to result in the same distribution; at the same time, the behavior of $f$ and $g$ may be different on those input values which do not occur as outputs of $p$.

  We finally introduce a notion of conditional independence, which is a straightforward generalization of~\cite[Defnition~12.12]{markov_cats} from the binary case to the $n$-ary case.

@@

Many theorems in probability theory involve infinitely many random variables at a time, together with a joint distribution for them. In the setting of Markov categories, this means that we need to consider morphisms whose codomain is an ``infinite tensor product'' in a suitable sense. In this section, we start by considering infinite tensor products in the more general context of semicartesian symmetric monoidal categories.
